Left-linear rewriting in adhesive categories

### 2024-04-21 concur/fusione.tex
concur branch

\title{Left-linear rules in adhesive categories}

@@

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  More precisely, we focus on left-linear rules, which
  not only consume, preserve and generate entities, as linear rules
  do, but may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules that have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

@@

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules a number of fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways. As we will observe, this prevents the development
of a sensible theory of concurrency since switches are used as a basis
for equating sequences of rewriting steps that differs only for the order of independent
steps.

@@

On the other hand, we show that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Specifically, 
the Church-Rosser theorem is fully recovered.
%for well-switching rewriting systems.

% \begin{enumerate}
% \item \emph{confluence of switching}: when transforming derivation
%   sequences into equivalent ones by switching independent steps, the
%   result not depends on the order in which switchings are applied, but
%   only on the resulting permutation;

% \item \emph{globality of independence}: if two independent steps are
%   moved forward or backward due to the application of switchings to a
%   derivation, they remain independent.
% \end{enumerate}

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key
to derive some basic but fundamental properties for linear rewriting
systems, namely

@@

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence via consistent permutations fails,
% : the existence of a consistent permutation is only necessary but
% not sufficient
hence it cannot be used for proving (1) and (2).
%
This leads to asking if the two properties above hold.
We show that much of the theory can be retained: globality of independence holds in a
weaker form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality, while consistency
of switching holds unchanged. Finally, we prove that a canonical form
for switching sequences can also be recovered.

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for future work.
The appendixes collect additional data, recalling basic results on
adhesive categories (Appendix~\ref{app:ade}), introducing and 
proving properties for \emph{strong enforcing} rewriting systems, a class of systems where the local Church-Rosser Theorem holds
(Appendix~\ref{app:fill}), and giving the proofs of the main 
results of our work (Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This first section recalls definitions and basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.

\begin{notation}
  We stipulate some notational conventions that will be used
  throughout the paper.
  \begin{itemize}
  \item
    Given a category $\X$ we will not distinguish notationally
    between $\X$ and its class of objects, so that ``$X\in \X$'' means
    that $X$ belongs to the class of objects of $\X$.

  %\item
    %If $1$ is a terminal object in $\X$, the unique
   % arrow $X\to 1$ from an object $X$ will be denoted by
    %$!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote
    %the unique arrow $0\to X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

  \item $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
    all arrows, monos and regular monos of $\X$, respectively.

  \item Given an integer $n\in \mathbb{Z}$, $[0,n]$ denotes the
    set of natural numbers less than or equal to $n$, assuming that
    $[0,n]=\emptyset$ whenever $n<0$.
  \end{itemize}
\end{notation}

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property that $\mathcal{M}$-adhesive categories enjoy is the so-called \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. We
will recall it and examine some of its consequences. First of all, we
need to introduce some terminology and facts regarding subclasses of
$\mor(\X)$.

\begin{definition}
  Let $\X$ be a category and $\mathcal{A}$ a subclass of
  $\mor(\X)$. We say that $\mathcal{A}$ is
  \begin{itemize}
    \parbox{11cm}{\item
      \emph{stable under pushouts (pullbacks)} if for every pushout (pullbacks) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
    \item \emph{closed under composition} if $g, f\in \mathcal{A}$ implies $g\circ f\in \mathcal{A}$ whenever $g$ and $f$ are composable.{\tiny }}\hfill
    \parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
    \parbox{11cm}{}\hfill
  \end{itemize}
\end{definition}

So equipped, we can introduce the notion of \emph{$\mathcal{A}$-Van Kampen square}.

\noindent
\parbox{10cm}{
  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider the diagrams aside.
 %   
%    \hspace{15pt}
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is an \emph{$\mathcal{A}$-Van Kampen square} if
    \begin{enumerate}
    \item it is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
    if and only if the front and right faces are pullbacks.
    \end{enumerate}
    Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

    We call a $\mor(\X)$-Van Kampen ($\mor(\X)$-stable) square simply a \emph{Van Kampen} (\emph{stable}) square.
  \end{definition}}
\parbox{2cm}{$\xymatrix@C=10pt@R=10pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D \\A \ar[dd]_{m}\ar[rr]^{f}&&B\ar[dd]^{n}\\ \\C\ar[rr]_{g} &&D}$ }

@@

%\todo{P: qui probabilmente si puo' compattare tenendo solo cio' che si usa}
%\todo{Si usa tutto per la classe B+ [DC]}
%$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
%\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]\label{lem:comma}
  %Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an
  %$\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a
  %functor that preserves $\mathcal{M}$-pushouts, and
  %$R:\B\rightarrow \C$ be a functor which preserves pullbacks.Then
 % $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where
  %\[
    %\cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in
    %\mathcal{M}, k\in \mathcal{M}'\}\]
%\end{theorem}
$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of slice and functor categories \cite{mac2013categories}, as shown by the following theorems from~\cite{ehrig2006fundamentals,lack2005adhesive}.

@@

\begin{theorem}\label{cor:slice}
  Let $X$ be an object of an $\mathcal{M}$-adhesive category
  $\X$. Then $\X/X$ and $X/\X$ are, respectively, $\mathcal{M}/X$-
  and $X/\mathcal{M}$-adhesive, where
  \[\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in \mathcal{M} \}
    \qquad X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in
    \mathcal{M} \}
  \]
\end{theorem}

\begin{theorem}
  \label{thm:functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Then for every small
  category $\Y$, the category $\X^\Y$ of functors $\Y\to \X$ is
  $\mathcal{M}^{\Y}$-adhesive, where
  \[\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}\]
\end{theorem}

@@

\begin{example}
  \label{ex:adhesive}
  Every topos is adhesive~\cite{lack2006toposes}, thus in particular
  $\cat{Set}$ is adhesive. By the closure properties recalled before,
  every presheaf $[\cat{X},\cat{Set}]$ is adhesive and in particular
  the category $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is
  adhesive where $E \rightrightarrows {V}$ is the two objects category
  with two morphisms $s,t \colon{E} \to {V}$. In a similar way,
  various categories of hypergraphs can be shown to be adhesive, 
  such as term graphs and hierarchical graphs~\cite{CastelnovoGM24}. We can also notice that the category $\cat{sGraphs}$ of simple graphs, i.e. graphs without
  parallel edges, is $\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not quasiadhesive.
\end{example}

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the Double-Pushout
 (DPO) approach. We will now recall the basic definitions and properties
of this approach. %to abstract rewriting.

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $\rho = (l,r)$ of
  arrows with $l: K \to L$ and $r: K \to R$ such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

  A \emph{left-linear rewriting system} is a pair $(\X, \R)$ where
  $\X$ is an $\mathcal{M}$-adhesive category and $\R$ a set of left
  $\mathcal{M}$-linear rules. The rewriting system $(\X, \R)$ is
  called $\mathcal{M}$-\emph{linear}
  %
  \full{\emph{(consuming)}}%
  %
  if every rule in $\R$ is
  $\mathcal{M}$-linear.
  \full{(consuming)}

@@

\begin{definition}[Derivation]
  Let  $\X$ be an $\mathcal{M}$-adhesive category and 
  $(\X, \R)$ a left-linear rewriting system. 
  Given $G, H \in \X$, a \emph{derivation}
  $\der{D}$ with source $G$ and target $H$, written
  $\der{D}: G \Mapsto H$, is defined as a sequence
  $\{\dder{D}_i\}_{i=0}^n$ of direct derivations such that
  \begin{enumerate}
  \item $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for
    every index $i$;
  \item $G_0=G$ and $G_{n+1}=H$.
  \end{enumerate}
  Moreover, we have an \emph{empty derivation} $G : G \Mapsto G$ for each $G \in \X$.
  %
  We denote by $\lgh(\der{D})$ the \emph{length} of derivation $\der{D}$.
 %
  \full{We will call the number $n+1$ the \emph{length} of the derivation,
  denoted by $\lgh(\der{D})$. We also say that an empty
  derivation has length $0$.}

@@

\noindent
\parbox{8cm}{
  \begin{definition}[abstraction equivalence]
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$ is a family of isomorphisms
    $\{\phi_X\}_{X\in \{G_i,D_i\}}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, and use the symbol $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R_i \ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}

>>>>

\title{Left-linear rewriting in adhesive categories}

@@

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exist for so-called linear rules which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm,
% i.e. where the behaviour is dictated by a set of rules that have to be instantiated with the system at hand,
concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

@@

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules a number of fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways.
%
As we will observe, this prevents the development of a sensible theory
of concurrency since switches are used as a basis for equating
sequences of rewriting steps that differs only for the order of
independent steps.

@@

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development has to face a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely

@@

The fact that for left-linear rewriting systems the characterisation
of switch equivalence via consistent permutations fails, and thus
% : the existence of a consistent permutation is only necessary but
% not sufficient
cannot be used for proving (1) and (2), leads to asking if these
properties hold.  We show that much of the theory can be
retained: globality of independence holds in a weaker form,
conceptually linked to the fact that fusions introduce a form of
disjunctive causality, while consistency of switching holds
unchanged. Finally, we prove that a canonical form for switching
sequences can also be recovered. These results represent a solid basis for
developing a satisfactory theory of concurrency for left-linear
rules.

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In we \S\ref{sec:equi} present the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class.
Finally, in \S\ref{conclusions} we outline possible directions for future work.
%
The appendices
% collect additional data,
recall basic results on adhesive categories (Appendix~\ref{app:ade}),
present \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}),
and give full proofs of the results
(Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This first section recalls definitions and basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so that
``$X\in \X$'' means that $X$ belongs to the class of objects of $\X$.
We will use $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$, assuming that $[0,n]=\emptyset$ whenever
$n<0$.

% \begin{notation}
%   We stipulate some notational conventions that will be used
%   throughout the paper.
%   \begin{itemize}
%   \item
%     Given a category $\X$ we will not distinguish notationally
%     between $\X$ and its class of objects, so that ``$X\in \X$'' means
%     that $X$ belongs to the class of objects of $\X$.

%   %\item
%     %If $1$ is a terminal object in $\X$, the unique
%    % arrow $X\to 1$ from an object $X$ will be denoted by
%     %$!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote
%     %the unique arrow $0\to X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

%   \item $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
%     all arrows, monos and regular monos of $\X$, respectively.

%   \item Given an integer $n\in \mathbb{Z}$, $[0,n]$ denotes the
%     set of natural numbers less than or equal to $n$, assuming that
%     $[0,n]=\emptyset$ whenever $n<0$.
%   \end{itemize}
% \end{notation}

% \begin{figure}
%   \centering
%   \begin{subfigure}[b]{0.45\textwidth}
%     \centering
%     \xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}        
%     \caption{}
%     \label{fig:vk:square}
%   \end{subfigure}
%   \hfill
%   \begin{subfigure}[b]{0.45\textwidth}
%     \centering
%     \xymatrix@C=10pt@R=10pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D}
%     \caption{}
%     \label{fig:vk:cube}
%   \end{subfigure}
%   \caption{Van Kampen property.}
%   \label{fig:vk}
% \end{figure}

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property that $\mathcal{M}$-adhesive categories enjoy is the so-called \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}. % We
% will recall it and examine some of its consequences.
In order to define it we need to introduce some terminology.
% and facts regarding subclasses of
% $\mor(\X)$.
%
%\begin{definition}
  Let  $\X$ be a category. A subclass $\mathcal{A}$ of
  $\mor(\X)$ is called
  \begin{itemize}
    \parbox{11cm}{\item
      \emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
    \item \emph{closed under composition} if $g, f\in \mathcal{A}$ implies $g\circ f\in \mathcal{A}$ whenever $g$ and $f$ are composable.{\tiny }}\hfill
    \parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
    \parbox{11cm}{}\hfill
  \end{itemize}
%\end{definition}

% So equipped, we can introduce the notion of \emph{$\mathcal{A}$-Van Kampen square}.

  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider the diagrams\\
    %
    \parbox{10cm}{
      aside.
      % 
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is an \emph{$\mathcal{A}$-Van Kampen square} if
      \begin{enumerate}
      \item it is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
      if and only if the front and right faces are pullbacks.
      \end{enumerate}}
    \parbox{2cm}{$\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }$ }\\
    %    
    Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

    We call a $\mor(\X)$-Van Kampen ($\mor(\X)$-stable) square
    % simply
    a \emph{Van Kampen} (\emph{stable}) square.
  \end{definition}

% old version
% \parbox{10cm}{
%   \begin{definition}[Van Kampen property]
%     Let $\X$ be a category and consider the diagrams aside.
%  %   
% %    \hspace{15pt}
%     Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
%     % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
%     is an \emph{$\mathcal{A}$-Van Kampen square} if
%     \begin{enumerate}
%     \item it is a pushout square;
%     \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
%     if and only if the front and right faces are pullbacks.
%     \end{enumerate}
%     Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

%     We call a $\mor(\X)$-Van Kampen ($\mor(\X)$-stable) square simply a \emph{Van Kampen} (\emph{stable}) square.
%   \end{definition}}
% \parbox{2cm}{$\xymatrix@C=10pt@R=10pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D \\A \ar[dd]_{m}\ar[rr]^{f}&&B\ar[dd]^{n}\\ \\C\ar[rr]_{g} &&D}$ }

@@

$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of slice and functor categories \cite{mac2013categories}, as shown by the following theorems from~\cite{ehrig2006fundamentals,lack2005adhesive}.

@@

\begin{theorem}
  \label{thm:slice-functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Given an object $X$
  the category $\X/X$ is $\mathcal{M}/X$-adhesive with
  $\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in
  \mathcal{M}$. Similarly, $X/\X$ is $X/\mathcal{M}$-adhesive with
  $X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in \mathcal{M}$.

  Moreover for every small category $\Y$, the category $\X^\Y$ of
  functors $\Y\to \X$ is $\mathcal{M}^{\Y}$-adhesive, where
  $\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}$.
\end{theorem}

@@

\begin{example}
  \label{ex:adhesive}
  $\cat{Set}$ is adhesive, and, more generally, every topos is
  adhesive~\cite{lack2006toposes}. By the closure properties above, every presheaf $[\cat{X},\cat{Set}]$ is adhesive, thus the category
  $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is adhesive
  where $E \rightrightarrows {V}$ is the two objects category with two
  morphisms $s,t \colon{E} \to {V}$. Similarly, various
  categories of hypergraphs can be shown to be adhesive, such as term
  graphs and hierarchical graphs~\cite{CastelnovoGM24}. Note that the category $\cat{sGraphs}$ of simple graphs, i.e.
  graphs without parallel edges, is
  $\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not
  quasiadhesive.
\end{example}

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the Double-Pushout
(DPO) approach.
%We next recall the basic definitions and properties of this approach.

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair  of
  arrows $\rho = (l: K \to L, r: K \to r)$ such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

  A \emph{left-linear rewriting system} is a pair $(\X, \R)$ where
  $\X$ is an $\mathcal{M}$-adhesive category and $\R$ a set of left
  $\mathcal{M}$-linear rules. It  is
  called $\mathcal{M}$-\emph{linear}
  %
  \full{\emph{(consuming)}}%
  %
  if every rule in $\R$ is
  $\mathcal{M}$-linear.
  \full{(consuming)}

@@

\begin{definition}[Derivation]
  Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ a
  left-linear rewriting system.  Given $G, H \in \X$, a
  \emph{derivation} $\der{D}$ with source $G$ and target $H$, written
  $\der{D}: G \Mapsto H$, is defined as a sequence
  $\{\dder{D}_i\}_{i=0}^n$ of direct derivations such that
  $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for every
  index $i$ and $G_0=G$, $G_{n+1}=H$.
  % \begin{enumerate}
  % \item $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for
  %   every index $i$;
  % \item $G_0=G$ and $G_{n+1}=H$.
  % \end{enumerate}
  Moreover, we have an \emph{empty derivation} $G : G \Mapsto G$ for each $G \in \X$.
  %
  We denote by $\lgh(\der{D})$ the \emph{length} of derivation $\der{D}$.
 %
  \full{We will call the number $n+1$ the \emph{length} of the derivation,
  denoted by $\lgh(\der{D})$. We also say that an empty
  derivation has length $0$.}

@@

\noindent
\parbox{8cm}{
  \begin{definition}[abstraction equivalence]
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear rewriting system. 
    An \emph{abstraction equivalence} between derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$ is a family of isomorphisms
    $\{\phi_X\}_{X\in \{G_i,D_i\}}$ such that the diagram on the right commutes
    for $i\in [0, \lgh(\der{D})]$. We say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, written $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R_i \ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}
